---
title: "NPC1 XGBoost Models"
author: "Adrian Castellanos"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
packages <- c("tidyverse", "magrittr", "sf", "terra", "xgboost", "caret", "readxl", "foreach", "SHAPforxgboost", "caret", "tidymodels", "ape", "phylotools", "patchwork")
sapply(packages, library, character.only = T)
setwd("~/Downloads/Filovirus/")
```

In this script, we find the general pipeline to run one of our XGBoost models of binding strength to a particular filovirus. Filoviruses can be changed out by find and replacing the label of them (e.g., from sudv to ebov or bomv, note that RAVV is listed as ravn here).

For the first part, we are going to read in our datasets compiling functional trait, phylogenetic, and habitat information. We have a distinct training dataset and our validation dataset will be subset from our prediction dataset. As with our data collation script, we use `case_when` to allow the viewer to see any taxonomic changes easily. We also apply a home range imputation based on methods from [Upham et al. 2023](https://doi.org/10.1101/504803). 
```{r readindata}
DATA <- read.csv("~/Downloads/Filovirus/npc1training_29Jul2025.csv")
VALD <- read.csv("~/Downloads/Filovirus/npc1prediction_29Jul2025.csv")
OBSV <- read.csv("bats_obs_pred.csv")
#MISM$species[which(is.na(match(MISM$species, ATRE$tip.label)))] 
OBSV %<>%
  mutate(species = case_when(species == "Unidentified Phyllostomid Bat" ~ NA,
                             species == "Unidentified Vespertillionid Bat" ~ NA,
                             species == "Scotophilus viridis nigritellus" ~ "Scotophilus viridis",
                             species == "Taphozous longimanes" ~ "Taphozous longimanus",
                             species == "Nycteris sp." ~ NA,
                             species == "Unidentified bat 1" ~ NA,
                             species == "Rhinolophus sp str1" ~ NA,
                             species == "Hipposideros ruber complex str2" ~ "Hipposideros ruber",
                             species == "Mormoops blainvillii" ~ "Mormoops blainvillei",
                             species == "Rhinolophus sp str 3" ~ NA,
                             species == "Rhinolophus sp str2" ~ NA,
                             species == "Artibeus glaucus" ~ "Dermanura glauca",
                             species == "Mops cf nanulus" ~ "Mops nanulus",
                             species == "Neoromicia nanus" ~ "Neoromicia nana",
                             species == "Chaerephon pumilus str1" ~ "Chaerephon pumilus",
                             species == "Chaerephon pumilus str2" ~ "Chaerephon pumilus",
                             species == "Eptesicus fuscus wetmorei" ~ "Eptesicus guadeloupensis", #in Eptesicus dutertreus which includes this taxon
                             species == "Hipposideros ruber complex str1" ~ "Hipposideros ruber",
                             species == "Megaderma sp" ~ NA,
                             species == "Myopterus daubentoni" ~ "Myopterus daubentonii",
                             species == "Pteronotus parnellii mesoamericanus" ~ "Pteronotus parnellii",
                             species == "Myonycteris angolensis smithii" ~ "Myonycteris angolensis",
                             species == "Unidentified Molossid Bat" ~ NA,
                             species == "Miniopterus fuliginosus" ~ "Miniopterus schreibersii",
                             species == "Pteropus dasymallus yayeyamae" ~ "Pteropus dasymallus",
                             species == "Pteropus sp" ~ NA,
                             species == "Vampyressa bidens" ~ "Vampyriscus bidens",
                             species == "Myonycteris angolensis" ~ "Lissonycteris angolensis",
                             species == "Myotis ricketti" ~ "Myotis pilosus",
                             species == "Mimon crenulatum" ~ "Gardnerycteris crenulatum",
                             T ~ species)) %>%
  filter(!is.na(species))
VALD <- VALD[VALD$binomial %in% setdiff(OBSV$species, DATA$species), ]
VALN <- VALD$binomial
DATA %<>%
   mutate(home_range_km2 = case_when(herbivore == 1 ~ 2.34 + 1.08 * log10(adult_mass_g),
                                    carnivore == 1 ~ 29.51 + 1.35 * log10(adult_mass_g),
                                    omnivore == 1 ~ 12.02 + 1.14 * log10(adult_mass_g)))
VALD %<>%
   mutate(home_range_km2 = case_when(herbivore == 1 ~ 2.34 + 1.08 * log10(adult_mass_g),
                                    carnivore == 1 ~ 29.51 + 1.35 * log10(adult_mass_g),
                                    omnivore == 1 ~ 12.02 + 1.14 * log10(adult_mass_g)))
```

This leaves us with our training dataset (`DATA`) and our separate validation dataset (`VALD`). For each, we will log transform heavily skewed variables, turn categorical variables into factors (more important for other algorithms, but retained here), and remove variables with near zero variation. 
```{r skewness}
SKEW <- sapply(c(16:138, 140:150), function(x) moments::skewness(na.omit(DATA[, x]))) #use moments package to determine skewness
SKEW

#wrap_plots(PLT[which(SKEW > 1.5)]) #only plot those with a skewness above 2
#Seems to be mostly temperature standard deviations and precipitation anomalies (along with percent crops)
CLSK <- colnames(DATA)[c(16:138, 140:150)][which(abs(SKEW) > 1.5)] #add heavily skewed variables to a list
#CLSK <- CLSK[c(1:19, 29, 32)]
#CLSK <- CLSK[c(1:8, 17, 34:35, 40:41)]
CLSK <- CLSK[c(1:8, 17, 34:35, 39:43)]


for(i in CLSK) {
  if(min(DATA[, i], na.rm = T) < 0) {
  DATA[, i] <- DATA[, i] + abs(min(DATA[, i], na.rm = T))
}
  if(min(DATA[, i], na.rm = T) == 0) {
  DATA[, i] <- DATA[, i] + 0.1
}
  DATA[, i] <- log(DATA[, i])
}
for(i in CLSK) {
  if(min(VALD[, i], na.rm = T) < 0) {
  VALD[, i] <- VALD[, i] + abs(min(VALD[, i], na.rm = T))
}
  if(min(VALD[, i], na.rm = T) == 0) {
  VALD[, i] <- VALD[, i] + 0.1
}
  VALD[, i] <- log(VALD[, i])
}
colnames(DATA)[colnames(DATA) %in% CLSK] <- paste("log", colnames(DATA)[colnames(DATA) %in% CLSK], sep = "_")
colnames(VALD)[colnames(VALD) %in% CLSK] <- paste("log", colnames(VALD)[colnames(VALD) %in% CLSK], sep = "_")

VARS <- colnames(DATA)[c(16:138, 140:150)]
DATA <- as.data.frame(DATA)
```

```{r categorical}
for(i in colnames(DATA)[c(36:37, 54:59, 63:65, 67:99, 104, 106:111, 134:136, 145)]) {
  DATA[, i] <- as.factor(as.character(DATA[, i]))
}
for(i in colnames(DATA)[c(36:37, 54:59, 63:65, 67:99, 104, 106:111, 134:136, 145)]) {
  VALD[, i] <- as.factor(as.character(VALD[, i]))
}
```

We will also remove all variables with a large amount of missing data, double check that we retained species names and family names for our records, and then create usable objects for XGBoost with the label and variables.
```{r nearzero}
NZV <- colnames(DATA)[c(16:138, 140:150)][caret::nearZeroVar(DATA[, c(16:138, 140:150)])] #just checking for variables with near zero variation
VARS <- VARS[!VARS %in% NZV]
VARS <- c(VARS[-which(apply(DATA[, VARS], 2, function(i) sum(is.na(i))) > 48)])
DATA[48, ]
SPECS <- c(DATA$species, VALD$binomial)
FAMS <- c(DATA$family, VALD$family)
FAMS[4] <- "Phyllostomidae"
VALD %<>%
  mutate(ebov_auc = OBSV[match(VALD$binomial, OBSV$species), "ebov_auc"])
DATA <- rbind(DATA[, c("ebov_auc", VARS)],
              VALD[, c("ebov_auc", VARS)])
alt <- apply(as.matrix(DATA[, c("ebov_auc", VARS)]), 2, as.numeric)
```

Given the smaller sample size, we use 100 bootstrap iterations to make our parameterization decisions and subsequent model evaluation. We are threading in the tidymodels framework to make parameterization and evaluation a bit easier. We restrict the trees to a low of 1,000 (to remove the very small number of trees) and a high of 10,000 (to remove the very high number of trees) and similarly restrict learning rate based on a range that worked best in preliminary runs. We use 4-fold cross validation to retain more data for testing. For each run, we save the full parameter combinations and their metrics, the chosen model's evaluation scores across all iterations, and, using SHAP values, importance and partial dependence data for each of the variables.
```{r xgboost}
DATA <- data.frame(alt)
set.seed(1876)
SEED <- sample(1:10000, 100)
xgb_spec <- boost_tree(trees = tune(),
                       tree_depth = tune(),
                       min_n = tune(),
                       loss_reduction =  tune(),
                       sample_size = tune(),
                       mtry = tune(),
                       learn_rate = tune()) %>%
  set_engine("xgboost", counts = FALSE) %>%
  set_mode("regression")
xgb_grid <- grid_space_filling(type = "latin_hypercube",
                               trees(c(1000, 10000)),
                                 tree_depth(),
                                 min_n(),
                                 loss_reduction(),
                                 sample_size = sample_prop(),
                                 mtry = finalize(mtry_prop(), TRN),
                                 learn_rate(range = c(0.001, 0.3), trans = NULL),
                                 size = 500)
OBJS <- foreach(i = 1:100) %do% {
  set.seed(SEED[i])
  SPL <- initial_split(DATA, prop = 0.75, breaks = 3, strata = ebov_auc)
  TRN <- training(SPL)
  TST <- testing(SPL)
  RECP <- recipe(ebov_auc ~ ., data = TRN)
  PREP <- prep(RECP)
  WRKF <- workflow() %>%
    add_formula(ebov_auc ~ .) %>%
    add_model(xgb_spec)
  FLDS <- vfold_cv(TRN,
                   breaks = 2,
                   v = 4,
                   strata = ebov_auc,
                   repeats = 1)
  list(SPL, WRKF, FLDS)
}
doParallel::registerDoParallel()
PARS <- foreach(i = 1:100, .combine = "rbind") %do% {
  xgb_res <- tune_grid(OBJS[[i]][[2]],
                       resamples = OBJS[[i]][[3]],
                       grid = xgb_grid,
                       control = control_grid(save_pred = F), 
                       metrics = metric_set(rmse, rsq, mae, huber_loss))
  xgb_res %>%
    collect_metrics()
}
PARS %>%
  group_by(mtry, trees, min_n, tree_depth, learn_rate, loss_reduction, sample_size, .metric) %>%
  summarise(mean_val = mean(mean, na.rm = T),
            n = n()) %>%
  ungroup() %>%
  pivot_wider(names_from = .metric,
              values_from = mean_val) %>%
  filter(!is.na(rmse),
         !is.na(rsq),
         !is.na(mae),
         !is.na(huber_loss)) %>%
  arrange(desc(rsq), rmse)
write.csv(PARS, "sudv_mammphylo_15Aug2025.csv", row.names = F)
PARS <- cbind.data.frame(mtry = 0.355, #914
                         min_n = 10, #12
                         tree_depth = 12, #11
                         learn_rate = 0.00561, #0.186
                         loss_reduction = 3.54, #0.0124
                         sample_size = 0.522, #954
                         trees = 1847) #7848
VALS <- foreach(i = 1:100, .combine = "rbind") %do% {
  FITS <- OBJS[[i]][[2]] %>%
    finalize_workflow(PARS) %>%
    last_fit(OBJS[[i]][[1]], metrics = metric_set(rmse, rsq, mae, huber_loss)) %>%
    collect_metrics %>%
    mutate(type = "test",
           std_err = NA) 
  TRNE <- OBJS[[i]][[2]] %>%
    finalize_workflow(PARS) %>%
    fit_resamples(OBJS[[i]][[3]], metrics = metric_set(rmse, rsq, mae, huber_loss)) %>%
    collect_metrics %>%
    mutate(type = "train") %>%
    dplyr::select(-n) %>%
    rename(.estimate = mean)
  rbind(FITS, TRNE)
}
VALS %>%
  group_by(.metric, type) %>%
  summarise(estimate = mean(.estimate, na.rm = T),
            sd_estimate = sd(.estimate, na.rm = T))
write.csv(VALS, "sudv_mammphylo_30Jul2025_values.csv", row.names = F)
SIMP <- foreach(i = 1:100, .combine = "rbind") %do% {
  xgb_fit <- OBJS[[i]][[2]] %>%
    finalize_workflow(PARS) %>%
    last_fit(OBJS[[i]][[1]], metrics = metric_set(rmse, rsq, mae, huber_loss)) %>%
    extract_fit_parsnip()
  TRN <- training(OBJS[[i]][[1]])
  shap.values(xgb_model = extract_fit_engine(xgb_fit),
              X_train = bake(PREP,
                             has_role("predictor"),
                             new_data = TRN,
                             composition = "matrix"))$mean_shap_score
}
write.csv(SIMP, "sudv_mammphylo_30Jul2025_simp.csv", row.names = F)

SHAP <- foreach(i = 1:100, .combine = "rbind") %do% {
  xgb_fit <- OBJS[[i]][[2]] %>%
    finalize_workflow(PARS) %>%
    last_fit(OBJS[[i]][[1]], metrics = metric_set(rmse, rsq, mae, huber_loss)) %>%
    extract_fit_parsnip()
  TRN <- training(OBJS[[i]][[1]])
  shap.prep(xgb_model = extract_fit_engine(xgb_fit),
            X_train = bake(PREP,
                           has_role("predictor"),
                           new_data = TRN,
                           composition = "matrix")) %>%
    mutate(iteration = i,
           row_index = OBJS[[i]][[1]]$in_id[.$ID])
}
write.csv(SHAP, "sudv_mammphylo_30Jul2025_shap.csv", row.names = F)
```

Using the full training dataset, we train an XGBoost model using the chosen parameters that gave the best performance and used the held out validation dataset to calculate a variety of performance metrics.
```{r}
MOD <- xgboost(data = as.matrix(DATA[, -1]),
                 label = DATA[, 1],
                 params = list(objective = "reg:squarederror",
                               eta = PARS$learn_rate,
                               max_depth = PARS$tree_depth,
                               subsample = PARS$sample_size,
                               colsample_bytree = PARS$mtry,
                               min_child_weight = PARS$min_n,
                               gamma = PARS$loss_reduction),
                 nrounds = PARS$trees,
                 verbose = F)
PVAL <- cbind.data.frame(OBSV[OBSV$species %in% VALN, c("species", "ebov_auc")]) %>%
  left_join(cbind.data.frame(species = VALN, pred_auc = predict(MOD, VALD)), by = "species")
ModelMetrics::rmse(PVAL$ebov_auc, PVAL$pred_auc)
ModelMetrics::rmse(PVAL$ebov_auc[c(-12, -15, -17, -37, -46)], PVAL$pred_auc[c(-12, -15, -17, -37, -46)])
```
